Tutorial: Quantifying Spatially Resolved Hydration Thermodynamics Using Grid Inhomogeneous Solvation Theory [Article v\versionnumber]

### 2024-06-28 manuscript/manuscript.tex
Updates from Overleaf

\usepackage{lipsum} % Required to insert dummy text
\usepackage[version=4]{mhchem}
\usepackage{siunitx}
\usepackage{float}
\usepackage{placeins}
\DeclareSIUnit\Molar{M}
\DeclareSIUnit\calorie{cal}
\DeclareSIUnit\kcalPerMolASqr{\kilo\calorie\per\mole\per\angstrom\squared}
%\DeclareSIUnit{\Molar}{M}
\DeclareSIUnit\angstrom{\text{\AA}}
%\DeclareSIPrefix\micro{\text{\textmu}}{0}
\sisetup{
	%math-rm=\mathsf,
	text-rm=\sffamily
}

@@

\author[1\authfn{1}]{Valentin J Hoerschinger}
\author[1\authfn{1}]{Franz Waibl}
\author[2]{Vjay Molino}
\author[2]{Helmut Carter}
\author[1]{Monica L Fern{\'a}ndez-Quintero}
\author[2]{Steven Ramsey}
\author[4]{Daniel R Roe}
\author[1*]{Klaus R Liedl}
\author[3*]{Michael K Gilson}
\author[2*]{Tom Kurtzman}

@@

\corr{klaus.liedl@uibk.ac.at}{KRL}
\corr{mgilson@health.ucsd.edu}{MKG}
\corr{thomas.kurtzman@lehman.cuny.edu}{TK}

%TODO Add ORCIDS
\orcid{Author 1 name}{AAAA-BBBB-CCCC-DDDD}
\orcid{Author 2 name}{EEEE-FFFF-GGGG-HHHH}

\contrib[\authfn{1}]{These authors contributed equally to this work}

@@

%%% Shortcuts and macros
\newcommand{\dgsolv}{\Delta G_\textup{solv}}
\newcommand{\software}{\texttt}
\newcommand{\code}{\texttt}
\newcommand{\todo}{\textcolor{red}}
\newcommand\inlinecode{\texttt}

@@

\begin{abstract}
Grid inhomogeneous solvation theory (GIST) is a method to compute the free energy of solvation of a solute molecule on a 3-dimensional grid based on molecular dynamics (MD) simulations.
The high spatial resolution of the GIST output, as well as the decomposition into energy and entropy contributions, allow for highly detailed analyses on both proteins and small molecules. However, this versatility also comes with a significant entry barrier for new users.

In this tutorial, we aim to guide the reader through the most common steps involved in a GIST analysis using the streptavidin-biotin complex as an demonstrative system. Towards this, Jupyter notebooks and Python packages are provided to simplify the analysis.
Furthermore, we discuss the theory of GIST with a focus on practical aspects.
We highlight potential pitfalls and show how to avoid technical difficulties.
For this tutorial, familiarity with molecular dynamics simulations as well as the AmberTools package \cite{Case2023-ambertools} is assumed.

@@

\section{Introduction}
Solvation thermodynamics govern many biological and chemical processes involving, e.g., aqueous solutions, hydrophobic effects, or partitioning between different environments.

Interactions with water are of special importance as many biological processes occur in aqueous environment \cite{Privalov2017-water-review}.
The free energy of solvation is defined as the reversible work needed to insert a solute molecule into the solvent from gas phase \cite{ben-naim-book}.
It is directly linked to the solubility of gases and drives many real-world effects such as hydrophobicity, ligand binding, and protein folding.

Grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012} is a method to compute the free energy of solvation, with applications in the context of drug design and biophysical property description.
It computes the solvent density from an MD trajectory and estimates localized entropic and enthalpic contributions on a grid around the solute molecule.
Since these contributions are spatially resolved, GIST improves interpretability and offers more detailed insights into the solvation of the investigated solute. 
The entropy is computed from the density distribution of the solvent around the solute molecule, in a first-order approximation. A schematic of this is shown in Figure~\ref{fig:carbazole}. 
The calculation of higher-order entropy terms is computationally demanding but was realized in recent works \cite{Nguyen2016-gist-second-order, Waibl2021-gist-salt}.
Nevertheless, solvation free energies of small molecule solutes can be computed at high accuracy using linear scaling factors for the higher-order terms, such as $\Delta S = 0.6\, \Delta S^\text{1st-order}$ for water \cite{Chen2021,Waibl2022-gist-solvents}.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/carbazole-figure.png}
    \caption{Schematic showing regions of high water density around a carbazole solute molecule. The energy is computed from the intermolecular interactions of a force field, while the entropy is approximated from the 3-dimensional solvent distribution around the solute molecule.}
    \label{fig:carbazole}
\end{figure}

Since its initial development, GIST has been used to describe the thermodynamics of water displacement in protein binding pockets \cite{Nguyen2016-gist-second-order,Ramsey2016,Balius2017-gist-ligand-discovery,Hufner-Wulsdorf2020-gist-drug-design}.
It has been used for a large-scale investigation of the hydration properties of proteins found in SARS-CoV-2 \cite{Olson2020-covid-gist}.
Furthermore, recent works used GIST to describe biophysical properties of serine proteases \cite{Kraml2019-gigist}, macrocycles \cite{Kamenik2020-gist-macrocycles}, or antibody fragements \cite{Waibl2021-gist-antibodies}.

@@

In this tutorial, we aim to give the reader an introduction to GIST and its application to calculate solution thermodynamics properties of interest. 
While previous publications \cite{Ramsey2016} and tutorials \cite{amber_tut_gist} treated how to setup and run GIST calculations in detail, the post-processing was often treated less rigorously. While building on these tutorials, we provide an updated guide to calculating solvation properties using GIST to address recent improvements.
Furthermore, we aim to establish best practices for the analysis and interpretation of GIST calculations.

\subsection{Comparison to other methods}
A wide range of methods have been devised to compute the free energy of hydration.
Data-driven methods include QSPR and UNIFAC \cite{Borhani2019-qspr,Fredenslund1975-unifac}.
Polarizable continuum models (PCMs) \cite{Miertus1981-pcm,Klamt1993-cosmo} treat the solvent as a homogeneous phase that reacts to the electrostatic potential of the solute molecule \cite{Mennucci2010-pcm}.
%Molecular mechanics based approaches describe the underlying PES through empirically fit force-fields, which reduce the computational demand of these calculations tremendously.
%This allows for the treatment of large biological systems such as proteins in aqueous solution.
Methods such as MM/PBSA or MM/GBSA \cite{Sitkoff1994-pbsa,Kollman2000-mmpbsa} combine an implicit treatment of solvent electrostatics with structural sampling of the solute molecule.
They can quickly provide results for a large number of structures \cite{Genheden2015-mmpbsa-review}.

@@

Methods derived from statistical mechanics bridge this gap by providing three-dimensional resolution.
These methods most commonly are based on either the Ornstein–Zernike (OZ) equation, or \cite{Hansen2013-simple-liquids} inhomogeneous fluid solvation theory (IST) \cite{Lazaridis1998}.
The most important example of OZ-based methods is the reference interaction site model (RISM) \cite{Chandler1972-rism} as well as its extension to three dimensions (3D-RISM) \cite{Kovalenko1998-3drism}.
Those methods compute the atomic solvent distributions by self-consistently solving the OZ integral equation to predict solvation thermodynamics.

On the other hand, methods based on IST generally use a molecular solvent distribution obtained from an MD simulation to compute the free energy of hydration from its enthalpic and entropic contributions, providing easier interpretability of the resulting distributions.
Examples include grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012,Ramsey2016}, WaterMap \cite{Young2007-watermap,Abel2008-watermap}, SSTMap \cite{Haider2018-sstmap}, STOW \cite{Li2012-stow} and Solvaware \cite{Huggins2016-solvaware}.
The main limitation of those methods is that the entropy is approximated through an infinite expansion in terms of the density distribution. 
Within GIST and other more recent methods, these entropy contributions are calulated via a k-nearest-neighbor (kNN) approach.  
Due to the high computational effort required to converge the higher order terms, the entropy expansion is usually truncated after the first or, sometimes, the second \cite{Nguyen2016-gist-second-order,Waibl2022-gist-solvents} term, thereby introducing an error.
While there are approximations available to deal with those problems, most noteworthy the mutual information expansion (MIE) and Maximum Information Spanning Tree (MIST) approaches, they are not currently implemented in GIST. 
Recent kNN implementations to calculate solvation entropy further improve phase space sampling by relabeling the solvent molecules in the simulation, treating the solvent molecules as distinguishable \cite{reinhard2007estimation,heinz2019computing,heinz2021permut,fogolari2021optimal}.For a recent review of kNN methods to calculate entropy in a molecular context, see \cite{fogolari2024k}.\\
Furthermore, IST does not consider the internal degrees of freedom of the solvent.
Therefore, IST-based approaches can only be used with sufficiently rigid solvents.
Nevertheless, they have been applied on a wide variety of problems involving both small molecules and biological systems as solute molecules.\\
GIST is an implementation of IST on a three-dimensional grid. Unlike the related method WaterMap \cite{Abel2008-watermap,Young2007-watermap}, it provides solvation thermodynamics on an entire grid, rather than only high-density clusters in a binding pocket.
This implies a higher versatility, but also puts more responsibility on the user in terms of correct post-processing.
Since the density is computed from an explicit solvent MD simulation, GIST provides an accurate picture of solvation at an atomistic level, even for demanding systems such as drug binding sites in proteins.
While the need for MD simulations comes at a large computational cost, GIST is still more efficient than end-state approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
On the other hand, the first-order approximation means that the solvation entropy is less accurately described than in TI or FEP \cite{Chen2021,Waibl2021-gist-salt}. 
Similar to the self-consistent integral theories, however, GIST only considers a single solute conformation and  multiple computations of representative structures are necessary to describe ensemble properties.

\subsection{Scope}
The purpose of this tutorial is to help users run GIST calculations and interpret the results.
The tutorial is structured around two examples: a GIST calculation for the small molecule biotin, and a GIST study of biotin-streptavidin binding.
We provide a workflow that can be easily adapted to different systems, an introduction to the method and the theory behind it, and an overview of the different implementations of GIST.

We start with a basic GIST analysis of biotin.
Building on the basics, we present a more advanced GIST study of the streptavidin-biotin complex, reflecting on many common aspects of GIST studies.
We show how to interpret the three-dimensional contributions of solvation free energy in a binding pocket and around the ligand.
Additionally, we compute the binding contribution of the free energy of solvation, which requires accurate post-processing of the GIST outputs to avoid unfavorable summation of bias.
We believe that this example study covers a wide range of applications.\\
Furthermore, we discuss several technical aspects of GIST studies, such as the normalization of voxel values and how to avoid double counting of energy contributions. 
Additionally, we provide a Python library (\software{gisttools}) to unify the analysis of GIST data produced by various versions of GIST and make the post-processing of GIST results more accessible.
After completing this tutorial, we expect the reader to be able to run their own GIST study.

@@

\subsection{Background knowledge}
This tutorial is aimed at users with a solid knowledge of molecular dynamics (MD) simulations.\\
The user should be able to run MD simulations and analyze the resulting trajectories. 
The MD simulations presented in this tutorial are run with the Amber simulation engine and analyzed using AmberTools \software{cpptraj}. 
For a conceptual overview of running MD simulations with Amber, we recommend reading the Amber manual (\url{https://ambermd.org/Manuals.php}) \cite{amber24} and the Amber tutorials (\url{https://ambermd.org/Manuals.php}). 
Previous experience of GIST is not necessary, although we recommend reading the works referenced in section~\ref{sec:theory}.

@@

\subsection{Software/system requirements}
%Tutorials should clearly define what system and/or software requirements the researcher will need to complete the tutorial (e.g., VMD version 1.9 or newer, AMBER, etc.). Tutorials requiring specific software packages must provide instructions and files for the referenced version of the software.
The main implementation of GIST is part of the MD analysis software \software{cpptraj} released with AmberTools \cite{amber24, Case2023-ambertools}.
A recent version of \software{cpptraj} should be used. We recommend using at least version 6.15.2, which can be obtained from GitHub (\url{https://github.com/Amber-MD/cpptraj}).

The Amber simulation engine is used in this tutorial to run the MD simulations.
While Amber is free for non-commercial use, it might not always be available in industrial settings.
In that case, other MD engines can be used as long as topologies and trajectories can be produced in a format supported by \software{cpptraj}.
If using GROMACS \cite{Abraham2015-gromacs,Pronk2013-gromacs}, we propose to prepare the topology with AmberTools \cite{Case2023-ambertools} and convert it using \software{acpype.py} \cite{Sousa_da_Silva2012-acpype}.
%We should actually test that!
Also note that compressed trajectory formats such as \inlinecode{xtc} might bias the entropy calculation due to the loss of precision for the atom positions. For further reference, the influence of compression on energy calculations was discussed in a recent publication \cite{Roe2022-compression}.

Furthermore, a recent Python version (3.6 or newer) should be available with the following packages installed:
\begin{itemize}
	\item \software{gisttools} (\url{https://github.com/liedllab/gisttools}, $\geq$ 0.2)
	\item \software{mdtraj} (\url{https://www.mdtraj.org/}, $\geq$ 1.9.7) \cite{McGibbon2015-mdtraj}
	\item \software{matplotlib} (\url{https://matplotlib.org/}, $\geq$ 3.7.0) \cite{Hunter2007-matplotlib}
	\item \software{numpy} (\url{https://numpy.org/}, $\geq$ 1.23.5) \cite{harris2020-numpy}
	\item \software{pandas} (\url{https://pandas.pydata.org/}, $\geq$ 1.5.3) \cite{pandas-2023_8239932}
\end{itemize}
If the reader prefers to skip the MD and GIST calculation, we also provide the GIST output files with this tutorial, such that the post-processing can be done without any expensive calculations.
We additionally provide a Jupyter notebook \cite{Kluyver2016-jupyter,Granger2021-jupyter} with the presented analyses. 
%We recommend JupyterLab to work with the notebook. Jupyter is available at \url{https://jupyter.org/}.

For the visualization of three-dimensional contributions we require molecular visualization software that can handle \software{OpenDX} files, such as \software{PyMol} \cite{pymol} or \software{VMD} \cite{vmd}.
Most of the visualisations and results in this tutorial can also be obtained using the \software{gistpp} program, which is available from \url{https://github.com/KurtzmanLab/Gist-Post-Processing} (code and documentation). \software{Gistpp} can directly manipulate \software{OpenDX} files from the GIST output. 
Its usage and some examples were shown in a previous tutorial on GIST \cite{Ramsey2016}.

\section{Running GIST}
\label{sec:running_GIST}
%\subsection{Running the GIST analysis}
To run a GIST analysis, the following is required:
\begin{itemize}
	\item \software{cpptraj}.
	\item A topology of the solute molecule of interest in a solvent box.
	\item An MD trajectory of that system.
\end{itemize}
Note that the solute should not move significantly for a GIST calculation, since IST does not take the solute degrees of freedom into account.
This is typically achieved by applying restraints on the solute heavy atoms during the MD simulation.
GIST analysis can be divided in three steps: defining a region of interest, running GIST using \software{cpptraj}, and post-processing and analysis.
\cite{Ramsey2016}

\paragraph{Identify and define the region of interest}
In order to run a GIST analysis, one must define the region where solvent properties will be calculated.
This region is defined by a rectangular grid with given position, size, and voxel spacing.
The default grid spacing of \SI{0.5}{\angstrom} provides reasonable discretization for many use-cases.
While integrals of GIST quantities over the grid are independent of the voxel size, provided sufficient sampling, individual values converge slower at high resolutions.
The grid is defined by parameters of the \inlinecode{gist} command in \software{cpptraj}.
The grid center is defined by \inlinecode{gridcntr}, the voxel number in x, y, and z direction by \inlinecode{griddim}, and the spacing (voxel side-length) in \AA{} by \inlinecode{gridspacn}:

\begin{lstlisting}[style=cpptraj]
gist gridcntr <x> <y> <z> \
griddim <Nx> <Ny> <Nz> gridspacn <val>
\end{lstlisting}
Additionally, the GIST calculation should usually be preceded by preprocessing steps, such as centering the solute molecule in the unit cell or reimaging to account for periodic boundary conditions.
Often, the region of interest will be defined as the surrounding of a molecule, which might for example be a ligand in the binding pocket or the whole protein. 
Using \inlinecode{cpptraj's } \inlinecode{bounds} command, the region of interest can be output to a file. For example, here is the command for a region of \SI{10}{\angstrom} around the ligand LIG:

\begin{lstlisting}[style=cpptraj]
bounds :LIG dx 0.5 offset 20 name Grid out bounds.dat
\end{lstlisting}
Note that the \inlinecode{bounds} command needs to be supplied with the exact same trajectory as used in the GIST calculation to generate correct boundaries. Therefore, any preprocessing steps need to match between the \inlinecode{bounds} command and the following \inlinecode{gist} calculation.

In addition, a helper script called \inlinecode{findcentroid.py} for the same purpose is supplied with this tutorial, and can be used as follows:

\begin{lstlisting}[style=bash]
python findcentroid.py -i structure.pdb
\end{lstlisting}

\paragraph{Run GIST using cpptraj}
GIST is implemented in \software{cpptraj} which requires the topology and the trajectory of the MD simulation to be analyzed.
The command needed to run GIST in cpptraj can be written in an input file (e.g gist.in) for convenience. The basic command, assuming a grid centered at coordinates (x=\SI{10.5}{\angstrom}, y=\SI{20.1}{\angstrom}, z=\SI{30.1}{\angstrom}) and a grid size of $30\times30\times30$ voxels with \SI{0.5}{\angstrom} spacing, is:

\begin{lstlisting}[style=cpptraj]
gist gridcntr 10.5 20.1 30.1 griddim 30 30 30 \
	gridspacn 0.50 out gist.dat
\end{lstlisting}
This command tells \software{cpptraj} to run a GIST calculation using the specified region and print the output in a file called \inlinecode{gist.dat}.  An overview of how to work with \software{cpptraj} is provided in the program's documentation \cite{cpptraj_doc}, AMBER tutorials \cite{amber_tut_cpptraj} and on the AMBER-hub website \cite{amber_hub}.

>>>>

\usepackage{lipsum} % Required to insert dummy text
\usepackage[version=4]{mhchem}
\usepackage{siunitx}
\DeclareSIUnit\Molar{M}
\DeclareSIUnit\calorie{cal}
\DeclareSIUnit\kcalPerMolASqr{\kilo\calorie\per\mole\per\angstrom\squared}
%\DeclareSIUnit{\Molar}{M}
\DeclareSIUnit\angstrom{\text{\AA}}
%\DeclareSIPrefix\micro{\text{\textmu}}{0}
\sisetup{
	%math-rm=\mathsf,
	text-rm=\sffamily
}

@@

\author[1\authfn{1}]{Franz Waibl}
\author[1\authfn{1}]{Valentin J Hoerschinger}
\author[2]{Vjay Molino}
\author[1]{Monica L Fern{\'a}ndez-Quintero}
%\author[1,2]{Firstname Middlename Familyname}
%\author[2]{Firstname Initials Surname}
\author[4]{Daniel R Roe}
\author[1*]{Klaus R Liedl}
\author[3*]{Michael K Gilson}
\author[2*]{Tom Kurtzman}

@@

\corr{klaus.liedl@uibk.ac.at}{KRL}  % Correspondence emails.  FMS and FS are the appropriate authors initials.
\corr{mgilson@health.ucsd.edu}{MKG}
\corr{thomas.kurtzman@lehman.cuny.edu}{TK}

\orcid{Author 1 name}{AAAA-BBBB-CCCC-DDDD}
\orcid{Author 2 name}{EEEE-FFFF-GGGG-HHHH}

\contrib[\authfn{1}]{These authors contributed equally to this work}
%\contrib[\authfn{2}]{These authors also contributed equally to this work}

%\presentadd[\authfn{3}]{Department of General, Inorganic and Theoretical Chemistry, University of Innsbruck, Austria}
%\presentadd[\authfn{4}]{Department, Institute, Country}

@@

%%% Shortcuts and macros
\newcommand{\dgsolv}{\Delta G_\textup{solv}}
\newcommand{\software}{\texttt}
\newcommand{\todo}{\textcolor{red}}
\newcommand\inlinecode{\texttt}

@@

\begin{abstract}
Grid inhomogeneous solvation theory (GIST) is a method to compute the free energy of solvation of a compound on a 3-dimensional grid based on molecular dynamics (MD) simulations.
The high spatial resolution of the GIST output, as well as the decomposition into energy and entropy contributions, allow for highly detailed analyses on both proteins and small molecules. However, this versatility also comes with a significant entry barrier for new users.

In this tutorial, we aim to guide the reader through the most common steps involved in a GIST analysis at the example of the streptavidin-biotin complex.
Furthermore, we discuss the theory of GIST with a focus on practical aspects.
We highlight potential pitfalls and show how to avoid technical difficulties.
We assume familiarity with molecular dynamics simulations as well as the AmberTools package.

@@

\section{Introduction}
Solvation thermodynamics govern many biological and chemical processes involving, e.g., aqueous solutions, hydrophobic effects, or partitioning between different environments.
Interaction with water are of special importance as many biological processes occur in aqueous environment \cite{Privalov2017-water-review}.
The free energy of solvation is defined as the reversible work needed to insert a solute into the solvent \cite{ben-naim-book}.
It is directly linked to the solubility of gases and drives many real-world effects such as hydrophobicity, ligand binding, and protein folding.

Grid inhomogeneous solvation theory (GIST) is a method to compute the free energy of solvation, with applications in the context of drug design and biophysical property description.
It estimates the solvent density from an MD trajectory and computes localized entropic and enthalpic contributions on a grid around the solute.
The entropy is computed from the density distribution of the solvent around the solute, in a first-order approximation. A schematic of this is shown in Figure~\ref{fig:carbazole}. Nevertheless, solvation free energies of small molecules can be computed at high accuracy using linear scaling factors for the higher-order terms, such as $\Delta S = 0.6\, \Delta S^\text{1st-order}$ for water \cite{Chen2021,Waibl2022-gist-solvents}.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/carbazole-figure.png}
    \caption{Schematic showing regions of high water density around a carbazole molecule. The energy is computed from the intermolecular interactions of a force field, while the entropy is approximated from the 3-dimensional solvent distribution around the solute.}
    \label{fig:carbazole}
\end{figure}

Since its initial development, GIST has been used to describe the thermodynamics of water displacement in protein binding pockets \cite{Nguyen2016-gist-second-order,Ramsey2016,Balius2017-gist-ligand-discovery,Hufner-Wulsdorf2020-gist-drug-design}.
It has also been deployed in systematically mapping the hydration properties of proteins found in SARS-CoV-2 \cite{Olson2020-covid-gist}.
Furthermore, recent works used GIST to describe biophysical properties of serine proteases \cite{Kraml2019-gigist}, macrocycles \cite{Kamenik2020-gist-macrocycles}, or antibody fragements \cite{Waibl2021-gist-antibodies}.

@@

In this tutorial, we aim to give the reader an introduction into GIST and its application to calculate solution thermodynamics properties of interest. 
While previous publications \cite{Ramsey2016} and tutorials\footnote{See \url{http://ambermd.org/tutorials/advanced/tutorial25/}} gave insightful pointers how to setup and run GIST calculations, the post-processing was often treated less rigorously. While building on these tutorials, we provide an updated guide to calculating solvation properties using GIST to address recent improvements.
Furthermore, we aim to establish a template for the analysis and interpretation of GIST calculations.

\subsection{Comparison to other methods}
A wide range of methods have been devised to compute the free energy of hydration.
Data-driven methods include QSPR and UNIFAC \cite{Borhani2019-qspr,Fredenslund1975-unifac}.
Polarizable continuum models (PCMs) \cite{Miertus1981-pcm,Klamt1993-cosmo} treat the solvent as a homogeneous phase that reacts to the electrostatic potential of the solute \cite{Mennucci2010-pcm}.
%Molecular mechanics based approaches describe the underlying PES through empirically fit force-fields, which reduce the computational demand of these calculations tremendously.
%This allows for the treatment of large biological systems such as proteins in aqueous solution.
Methods such as MM/PBSA or MM/GBSA \cite{Sitkoff1994-pbsa,Kollman2000-mmpbsa} combine an implicit treatment of solvent electrostatics with structural sampling of the solute.
They can quickly provide results for a large number of structures \cite{Genheden2015-mmpbsa-review}.

@@

Methods derived from statistical mechanics bridge this gap by providing three-dimensional resolution.
Two major flavors of methods in this group are those based on the Ornstein--Zernike (OZ) equation \cite{Hansen2013-simple-liquids} and those based on inhomogeneous fluid solvation theory (IST) \cite{Lazaridis1998}.
The most important example of OZ-based methods is the reference interaction site model (RISM) \cite{Chandler1972-rism} as well as its extension to three dimensions (3D-RISM) \cite{Kovalenko1998-3drism}.
Those methods compute the atomic solvent distributions by self-consistently solving the OZ integral equation to predict solvation thermodynamics.

On the other hand, methods based on IST generally use a molecular solvent distribution obtained from an MD simulation to compute the free energy of hydration from its enthalpic and entropic contributions.
Examples include grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012,Ramsey2016}, WaterMap \cite{Young2007-watermap,Abel2008-watermap}, SSTMap \cite{Haider2018-sstmap}, STOW \cite{Li2012-stow} and Solvaware \cite{Huggins2016-solvaware}.
The main limitation of those methods is that they rely on an infinite expansion of the entropy in terms of the density distribution, which is truncated after the first or, sometimes, the second \cite{Nguyen2016-gist-second-order,Waibl2022-gist-solvents} term.
Furthermore, IST does not consider the internal degrees of freedom of the solvent.
Therefore, IST-based approaches can only be used with sufficiently rigid solvents.
Nevertheless, they have been applied on a wide variety of problems involving both small molecules and biological systems.

GIST is an implementation of IST on a three-dimensional grid. It can be seen as a more general version of WaterMap \cite{Abel2008-watermap,Young2007-watermap} that provides solvation thermodynamics on the entire grid, rather than high-density clusters in a binding pocket.
This implies a higher versatility, but also puts more responsibility on the user in terms of correct post-processing.

Since the density is computed from an MD simulation, GIST can accurately treat confined water regions, which are challenging for implicit-solvent methods such as GBSA and self-consistent theories such as 3D-RISM \cite{Kovalenko1998-3drism}.
While the need for MD simulations comes at a large computational cost, GIST is still more efficient than end-state approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
On the other hand, the first-order approximation means that the solvation entropy is less accurately described than in TI or FEP \cite{Chen2021,Waibl2021-gist-salt}.
%The 3D-RISM approach allows for a decomposition into entropic and enthalpic terms as well as a full localization of the underlying distributions to the solute structure.
%Here you would explain what problem you are tackling and briefly motivate your work.
%
%In this particular template, we have removed most of the usage examples which occur in \texttt{sample-document.tex} to provide a minimal template you can modify; however, we retain a couple of examples illustrating more unusual features of our templates/article class, such as the checklists, and information on algorithms and pseudocode.
%
%Keep in mind, as you prepare your manuscript, that you should plan for a representative image  which will be used to highlight your article on the journal website and publications. Usually, this would be one of your figures, but it must also be uploaded separately upon article submission. We give specific guidelines for this image on the journal website in the section on article submission (see \url{https://livecomsjournal.github.io/authors/policies/index.html#article-submission}).
%
%Additionally, for well-formatted manuscripts, we recommend that you let LaTeX handle figure/table placement for you as much as possible, so please avoid specifying strenuous float instructions like `[h!]` and `[H]` as much as possible.

\subsection{Scope}
The purpose of this tutorial is to help users run GIST calculations and interpret the results.
The tutorial is structured around two examples: a GIST calculation for the small molecule biotin, and a study of biotin-streptavidin binding.
We provide a GIST workflow that can be easily adapted to different systems, an introduction to the method and the theory behind it, and an overview of the different implementations of GIST.

We start with a basic GIST analysis at the example of biotin.
Building on the basics, we present a more advanced GIST study of the streptavidin-biotin complex, reflecting on many common aspects of GIST studies.
We show how to interpret the three-dimensional contributions of solvation free energy in a binding pocket and around the ligand.
Additionally, we compute the binding contribution of the free energy of solvation, which requires accurate post-processing of the GIST outputs to avoid unfavorable summation of bias.
We believe that this example study covers a wide range of applications.

Furthermore, we discuss several technical aspects of GIST studies, such as the normalization of voxel values and how to avoid double counting of energy contributions. 
Additionally, we provide a Python library (\software{gisttools}) to unify the analysis of GIST data produced by various versions of GIST and make the post-processing of GIST results more accessible.

After completing this tutorial, we expect the reader to be able to run their own GIST study.

@@

\subsection{Background knowledge}
This tutorial is aimed at users with a solid knowledge of molecular dynamics (MD) simulations.
The user should be able to run MD simulations and analyze the resulting trajectories.
Previous experience of GIST is not necessary, although we recommend reading the works referenced in section~\ref{sec:theory}.

@@

\subsection{Software/system requirements}
%Tutorials should clearly define what system and/or software requirements the researcher will need to complete the tutorial (e.g., VMD version 1.9 or newer, AMBER, etc.). Tutorials requiring specific software packages must provide instructions and files for the referenced version of the software.
The main implementation of GIST is part of the MD analysis software \software{cpptraj} released with AmberTools \cite{amber22}.

A recent version of \software{cpptraj} should be used. We recommend using at least version 6.15.2, which can be obtained from GitHub (\url{https://github.com/Amber-MD/cpptraj}).

The Amber simulation engine is used in this tutorial to run the MD simulations.
While Amber is free for non-commercial use, it might not always be available in industry settings.
In that case, other MD engines can be used as long as topologies and trajectories can be produced in a format supported by \software{cpptraj}.
If using GROMACS \cite{Abraham2015-gromacs,Pronk2013-gromacs}, we propose to prepare the topology with AmberTools and convert it using \software{acpype.py} \cite{Sousa_da_Silva2012-acpype}.
%We should actually test that!
Also note that compressed trajectory formats such as \inlinecode{xtc} might bias the entropy calculation due to the loss of precision for the atom positions. For further reference, the influence of compression on energy calculations was discussed in a recent publication \cite{Roe2022-compression}.

Furthermore, a recent Python version (3.6 or newer) should be available with the following packages installed:
\begin{itemize}
	\item \software{gisttools} (\url{https://github.com/liedllab/gisttools}, $\geq$ 0.2)
	\item \software{mdtraj} (\url{https://www.mdtraj.org/}, $\geq$ 1.9.7) \cite{McGibbon2015-mdtraj}
\end{itemize}

If the reader prefers to skip the MD and GIST calculation, we also provide the GIST output files with this tutorial, such that the post-processing can be done without any expensive calculations.
We additionally provide a Jupyter notebook \cite{Kluyver2016-jupyter,Granger2021-jupyter} with the presented analyses. 
%We recommend JupyterLab to work with the notebook. Jupyter is available at \url{https://jupyter.org/}.

For the visualization of three-dimensional contributions we require molecular visualization software that can handle \software{OpenDX} files, such as \software{PyMol} \cite{pymol} or \software{VMD} \cite{vmd}.

Most of the visualisations and results in this tutorial can also be obtained using the \software{gistpp} program, which is available from \url{https://github.com/KurtzmanLab/Gist-Post-Processing} (code and documentation). \software{Gistpp} can directly manipulate \software{OpenDX} files from the GIST output. 
Its usage and some examples were shown in a previous tutorial on GIST\cite{Ramsey2016}.

\section{Running GIST}
\label{sec:running_GIST}
%\subsection{Running the GIST analysis}
To run a GIST analysis, the following is required:
\begin{itemize}
	\item \software{cpptraj}.
	\item A topology of the molecule of interest in a solvent box.
	\item An MD trajectory of that system.
\end{itemize}
GIST analysis can be divided in three steps: defining a region of interest, running GIST using \software{cpptraj}, and post-processing and analysis.
\cite{Ramsey2016}

\paragraph{Identify and define the region of interest}
In order to run a GIST analysis, one must define the region where solvent properties will be calculated.
This region is represented by a rectangular grid with given position, size, and voxel spacing.
The default value of \SI{0.5}{\angstrom} provides reasonable discretization for many use-cases.
While integrals of GIST quantities over the grid are independent of the voxel size, individual values converge slower at high resolutions.

The grid is defined by parameters of the \inlinecode{gist} command in \software{cpptraj}.
The grid center is defined by \inlinecode{gridcntr}, the voxel number in x, y, and z direction by \inlinecode{griddim}, and the spacing (voxel side-length) in \AA{} by \inlinecode{gridspacn}.

\begin{lstlisting}[style=cpptraj]
gist gridcntr <x> <y> <z> \
griddim <Nx> <Ny> <Nz> gridspacn <val>
\end{lstlisting}

Often, the region of interest will be defined as the surrounding of a molecule, which might for example be a ligand in the binding pocket or the whole protein. Using \inlinecode{cpptraj's } \inlinecode{bound} command, the region of interest can be output to a file. For example, here is the command for a region of 10\AA{} around the ligand LIG:
\begin{lstlisting}[style=cpptraj]
bounds :LIG dx 0.5 offset 20 name Grid out bounds.dat
\end{lstlisting}
Note that any loading and preprocessing, such as imaging, needs to be run before this command and needs to be the exact same as is used for the \inlinecode{gist} command.
Additionally, a helper script called \inlinecode{FindCentroid.py} for the same purpose is supplied with this tutorial, and can be used as follows:

\begin{lstlisting}[style=bash]
python FindCentroid.py -i structure.pdb
\end{lstlisting}

\paragraph{Run GIST using cpptraj}
GIST is implemented in \software{cpptraj} which requires the topology and the trajectory of the MD simulation to be analyzed.
The command needed to run GIST in cpptraj can be written in an input file (e.g gist.in) for convenience. The basic command, assuming a given grid center and a grid size of $30\times30\times30$ voxels with \SI{0.5}{\angstrom} spacing, is:

\begin{lstlisting}[style=cpptraj]
gist gridcntr 10.5 20.1 30.1 griddim 30 30 30 \
	gridspacn 0.50 out gist.dat
\end{lstlisting}

Note that the solute should not move significantly for a GIST calculation, since IST does not take the solute degrees of freedom into account.
This is typically achieved using restraints on the solute heavy atoms.
This command tells \software{cpptraj} to run a GIST calculation using the specified region and print the output in a file called \inlinecode{gist.dat}. You will need to load and image your trajectory before running this command. An overview of how to work with \software{cpptraj} is provided in the program's documentation \cite{cpptraj_doc}, AMBER tutorials \cite{amber_tut_cpptraj} or on the AMBER-hub website \cite{amber_hub}.

To use OpenMP or GPU-accelerated (using CUDA) versions of cpptraj, a different executable name such as \inlinecode{cpptraj.OMP} or \inlinecode{cpptraj.cuda} is often used. Therefore, make sure to use the right compilation and executable, to get full benefit of the various improvements to  GIST.
The \inlinecode{gist} action has a large number of options to modify the GIST calculation or its output.
A full list is provided at the end of this tutorial or through the \software{cpptraj} manual.
The most important options are summarized in the following table:
